---
title: "Accessing International Debt Statistics (IDS) through the World Bank Data API"
output: html_document
---
```{r global_options, include=FALSE}
#removing all warnings and messages from the doc that appear when loading the packages
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
## Part 1 - Query the API: indicators and locations
### *R*
Date: 05/22/2019 <br>
[Data source](https://datacatalog.worldbank.org/dataset/international-debt-statistics)<br>
[Developer Information](https://datahelpdesk.worldbank.org/knowledgebase/topics/125589-developer-information)

This is first part of a two-part series on accessing the International Debt Statistics (IDS) database through the World Bank Data API. In this part, we will query the World Bank Data API to retrieve indicator names and location codes. In [Part 2](https://worldbank.github.io/debt-data/api-guide/ids-api-guide-r-2.html), we will use that information to explore the **regional trends of long-term external debt stocks** from the IDS database. <br>
The following code in this guide will show step-by-step how to:

1. Setup up your environment with the needed R packages
2. Create Indicator API Queries
3. Create Location API Queries

## 1. Setup
To start, make sure you have the following packages installed on your machine. To install an R package, type install.packages("wbstats") with the correct package name into R. You can also visit each of the linked packages for reference.

- [wbstats](https://cran.r-project.org/web/packages/wbstats/vignettes/Using_the_wbstats_package.html): Searching and downloading data from the World Bank Data API
- [httr](https://cran.r-project.org/web/packages/httr/index.html): Submitting HTTP requests
- [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html): Parsing json response
- [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html):  Data work


Then, open up your preferred mode of writing R, like [R Studio](https://www.rstudio.com/). Now follow the rest of the steps below to query the World Bank Data API to find your indicator and location codes.

```{r}
library(wbstats)
library(httr)
library(jsonlite)
library(tidyverse)
```

## 2. Indicator API Queries
To get a data series from the World Bank Data's API, you first need to use an "indicator code." For example, the indicator code for long-term external debt stock is "DT.DOD.DLXF.CD." These indicator codes can be found in the [World Bank Data Catalog](https://datacatalog.worldbank.org/), but you can also use the API to explore and select indicators. In this section, we will guide you through going from a topic idea to using the World Bank API to select an indictor code.

More information on this can also be found through the [Data Help Desk](https://datahelpdesk.worldbank.org/knowledgebase/articles/898599-indicator-api-queries).

First, we should choose a topic. Topics are high level categories to which all indicators are mapped, such as Education, Trade, etc. The topic in this guide is "External Debt," but we will need the topic's ID in order to see all its corresponding indicator names and codes.

#### Getting the Topic ID for Debt

The following GET request will get us every topic from the World Bank Data API. However, the request is returned in a json format that is difficult to read as is, so we will include some code to parse through it. Then we can see each topic name and ID available in the World Bank Data API.

```{r}
# Get all the topic names from the World Bank API
topicRequest <- GET(url = "http://api.worldbank.org/v2/topic?format=json")
topicResponse <- content(topicRequest, as = "text", encoding = "UTF-8")

# Parse the JSON content and convert it to a data frame. 
topicsJSON <- fromJSON(topicResponse, flatten = TRUE) %>%
  data.frame()

# Print the data frame and review the topic names and ids
cols <- c("id","value")
topicsJSON[,cols]
```
We can now see that "External Debt" topic has the ID "20"

#### Getting the Indicator code

Now that we have the ID for the topic "External Debt," we can make another request to the World Bank API to receive all of the indicator names and codes associated with that topic. In the code below, notice that the ID for "External Debt" is placed in the URL between "topic" and "indicator."

Most likely, there will be more External Debt indicators available than can fit on one page, the default amount is 50 results. So we will set the "per_page" parameter to 250. This will allow us to view all the results with one query, instead of having to request multiple pages. Then we can parse through the result to explore the different indicator Names and corresponding IDs. Because there so many indicators, the results are also saved as "ids-indicators.csv." (LINK NEEDED)

```{r}
# make request to World Bank API
indicatorRequest <- GET(url = "http://api.worldbank.org/v2/topic/20/indicator?per_page=250&format=json")
indicatorResponse <- content(indicatorRequest, as = "text", encoding = "UTF-8")

# Parse the JSON content and convert it to a data frame. 
indicatorsJSON <- fromJSON(indicatorResponse, flatten = TRUE) %>%
  data.frame()

# Print the data frame and review the indicator names and ids
cols <- c("id","name")
#indicatorsJSON[,cols] # to view all the indicator names, remove the # at the beginning of this line
```


The resulting list gives us all of the External Debt indicators and their codes. The indicator and code we want is "DT.DOD.DLXF.CD External debt stocks, long-term (DOD, current US$)." Before using this data, we need to understand its full definition. You can use an API query to get that information as well.

```{r}
# use the indicator code to define x
x <- "DT.DOD.DLXF.CD"


# use the above indicator code to find associate "sourceNote" or definition
y <- which(indicatorsJSON$id == x)
print(indicatorsJSON$sourceNote[y])

```

By using the API queries above, we were able sort through the World Bank API topics, find "External Debt," explore the corresponding indicators, and then select one indicator of interest and find its ID and definition.

## 3. Location API Queries

Now that we have the indicator code, we need to select a location. To select a location by country, region, or income level category you will need to know its 2 or 3 letter code. This section will show you how to use similar API queries to pull the location names and codes. 

For more information on this visit the [Data Help Desk](https://datahelpdesk.worldbank.org/knowledgebase/articles/898590-country-api-queries).

When pulling the list of locations, I will set the per_page result to 300, as there will too many results to display on one page.

```{r}
# Make request to World Bank API
locationRequest <- GET(url = "http://api.worldbank.org/v2/country?per_page=300&format=json")
locationResponse <- content(locationRequest, as = "text", encoding = "UTF-8")

# Parse the JSON content and convert it to a data frame. 
locationsJSON <- fromJSON(locationResponse, flatten = TRUE) %>%
  data.frame()
```

From this list of locations, we need to pick one or more locations and then find the corresponding IDs and names Because there are so many locations, the results are also saved as [location-codes csv](https://github.com/worldbank/debt-data/blob/master/location-codes.csv)*.

*The location-codes csv was created using the API query: http://api.worldbank.org/v2/sources/2/country/data
```{r}
# Createa list of the location IDs and names
cols <- c("id","name")
locationList <- locationsJSON[,cols]
```

From the resulting list, you can select the location IDs for the countries or regions of interest.

#### Now head to Part 2 of this series to find out how to use the indicator and location codes to call a data series from the World Bank Data API
### [Part 2 - Get and explore the data](https://worldbank.github.io/debt-data/api-guide/ids-api-guide-r-2.html)











